---
title: "Gene Signature for Overall Surival with TCGA-PAAD Methylation Dataset (first preference)"
output: html_document
author:
- name: Aditya Pawar

date: "2024-06-12"
---

#### This script uses the DNA-methylation(PAAD) and PAAD clinical data from TCGA to analyse survival of the patients. We are using Elastic net multivariate anlaysis to calculate survival and plot Kaplan Meier plots based on Dr. Manne's gene panel (first preference). The alpha used for this script is calculated based on the lowest mean error and the minimum lambda value accordinly for every iteration(100 iterations to calculate best value for alpha and lambda). We have done 100 different iterations.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library,  warning=FALSE,message=FALSE}
library(dplyr)
library(survival)
library(survminer)
library(glmnet)
```

#### Load and clean the data

```{r data}
load("PAAD_meth.RData")

# names(PAAD_meth)
# Rename specific columns
names(PAAD_meth)[names(PAAD_meth) == "Row.names"] <- "cpg_reg"
names(PAAD_meth)[names(PAAD_meth) == "UCSC_RefGene_Name"] <- "genes"
names(PAAD_meth)[names(PAAD_meth) == "UCSC_RefGene_Group"] <- "group"
names(PAAD_meth)[names(PAAD_meth) == "Relation_to_UCSC_CpG_Island"] <- "island"

# Remove rows where the name is "Unknown"
PAAD_meth <- PAAD_meth[PAAD_meth$genes != "Unknown", ]

# Combine the two columns with a space in between
PAAD_meth$cpg_genes <- paste0(PAAD_meth$cpg_reg,"_", PAAD_meth$genes)

genes_PAAD <- sort(c("BNIP3", "CADM1", "CLDN5", "DISP3", "EBF3", "EYA4", "FOXE1", "NXPH1", "RELN", "SLIT2", "SOX17", "ADAMTS1", "APC", "BMP3", "BNC1", "BRCA1", "CCDC181", "CCND2", "CDKN1C", "CDKN2B", "DAPK1", "EPB41L3", "ESR1", "ALKAL1", "GATA4", "GATA6", "GPC3", "HOXA1", "HOXA7", "IGF1", "LRFN5", "MAPT", "MARCHF11", "MEIS2", "MEST", "MGMT", "MIR-129-2/129", "MLH1", "MUC2", "MYOD1", "NPTX2", "ONECUT1", "CDKN2A", "PANTR1", "PCDH10", "PENK", "PGK1", "PGR", "PLAU", "PRKCB", "PROX1", "PXDN", "RARB", "RASSF1", "RB1", "RUNX3", "SEPTIN9", "SFRP1", "SFRP2", "SIX3", "SLC9A3", "SOCS1", "SPARC", "SPSB4", "CD2", "SST", "SYK", "TBR1", "TEAD1", "THBS1", "TYMS", "TRIM73", "UCHL1", "VHL", "VWC2", "YAP1", "ZNF781", "SEMA5A", "TFPI2", "MIR-1292/3P", "MIR-1292/5P", "MUC5AC", "ADAMTS1", "ACIN1", "ALX4", "BMP3", "BNC1", "CDKN2A", "GSTP1", "HIC1", "MSX2", "MEST", "MGMT", "MLH1", "NEUROG1", "NPTX2", "PCDH10", "PENK", "POU4F1", "SEMA5A", "SEPTIN9", "SFRP1", "SFRP2", "SPSB4", "TFPI2", "TNFRSF10C", "WNT5A", "UCHL1", "GSTM1", "ONECUT2", "ASXL2", "TET1", "TSPAN1", "ABCC1", "ABCG2", "CDA", "ABCB11", "ABCB4", "ABCC10", "ABCC3", "METTL3", "DDX3X", "SLC29A1", "CADM1", "PTGES2", "HSPA5", "SNRPF", "ABCC5", "CES2", "SPARC", "BRCA1", "EGFR", "TNFSF10", "ABCB1", "MYEOV", "HMGA1", "JAG1", "LCN2", "NOTCH2", "RRM2", "TM4SF1", "RRM1", "MUC5AC", "SLC29A3", "SFN", "DPYD", "MAP2", "MUTYH", "SLC28A1", "BCL2L1", "MAP2", "ABCC2", "SLC22A3", "MLH1", "SFN", "VASH2", "BNIP3", "SST", "DKK3", "ARID1A", "GSTM2", "ISG15", "SMARCA2", "MUC4", "HNF1A", "NT5C1A", "TYMS", "ABCB5", "ABCC11", "SLC28A2", "SLC22A7", "ZEB1", "TFAP2E", "NRP1", "SLC28A3", "MCL1", "PYCARD", "MUTYH", "SOX8", "CHFR", "SLFN11", "MAP3K7", "CXCL8", "SLFN11", "SLC22A2", "SLFN11", "ZEB1", "CTNNB1", "IGFBP3", "PROKR2", "BNIP3", "ABCC1", "ABCC2", "ABCC3", "ABCC4", "ABCC5", "ABCC6", "CFTR", "ABCC8", "ABCC9", "ABCC10", "ABCC11", "ABCC12", "ABCC13", "USP8", "PRMT1", "RUNX1", "UBE2T", "GSDME", "TGM2", "DCK", "TPX2", "CPT1B", "ENO1", "CPT1B", "ATM", "DHX38", "RELL2", "GLO1", "LDHA", "CTPS2", "BICC1", "HSPB1", "CASC9", "TACC3", "ITGA3", "CXCL5", "APOL1", "ANGPTL4", "SLC2A1", "ALDOB", "KCNH2", "ABCB1", "TFAM", "TMEM139", "RRP9", "CD44", "IGF1R", "HSPB1", "TRIM31", "YEATS4", "EIF5A", "CCDC85A", "TGM2", "ITGB4", "GSDMC", "SLC38A5"))

genes_PAAD <- unique(genes_PAAD)
gene.id <- PAAD_meth$genes
gene.num <- which(gene.id %in% genes_PAAD)
gene.nomatch <- genes_PAAD[-which(genes_PAAD %in% gene.id)]
gene.match <- genes_PAAD[which(genes_PAAD %in% gene.id)]
gene.match

gene.nomatch

## SElect only matched genes from DNA meth data
# Filter the data frame
PAAD_meth_match <- PAAD_meth %>%
  filter(genes %in% gene.match)


## transpose data for selected genes
columns_to_drop <- c("cpg_genes","cpg_reg","genes","group","island")
PAAD_meth_match_col <- PAAD_meth_match[, !(names(PAAD_meth_match) %in% columns_to_drop)]
PAAD_meth_match_mat <- as.data.frame(t(PAAD_meth_match_col))
colnames(PAAD_meth_match_mat) <- PAAD_meth_match$cpg_genes
PAAD_meth_match_mat <- cbind(sample = colnames(PAAD_meth_match_col), PAAD_meth_match_mat)

## Clean Clinical data
load("clinical_PAAD.Rdata")

## Combine survival and meth data
PAAD_meth_surv <- merge(clinical_PAAD,PAAD_meth_match_mat, by="sample")
columns_to_drop <- c("_PATIENT","sample","DSS","DSS.time","DFI","DFI.time","PFI","PFI.time","Redaction")
PAAD_meth_surv <- select(PAAD_meth_surv, -any_of(columns_to_drop))
PAAD_meth_surv$OS.time <- PAAD_meth_surv$OS.time/365

## Remove data with time == 0
PAAD_meth_surv <- PAAD_meth_surv[(!PAAD_meth_surv$OS.time == 0),]
# head(PAAD_meth_surv)
```

Define Variables
```{r Define variables}
p.logrank.all <- rep(2,100)
fit.all <- rep(2,100)
```

#### Kaplan Meier Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
for (i in 1:100) {
  print(paste0("iteration:",i))
  set.seed(i)
  X <- PAAD_meth_match_mat[,-1]
  time <- PAAD_meth_surv$OS.time
  # time[time == 0] <- 1
  status <- as.numeric(PAAD_meth_surv$OS)
  S <- Surv(time,status)
  K <- 3
  fold <- split(sample(1:nrow(X)), rep(1:K, length=nrow(X)))
  id.train <- c(fold[[1]],fold[[2]])
  id.test <- fold[[K]]
  X.train <- X[id.train,]
  S.train <- S[id.train,]
  time.train <- time[id.train]
  status.train <- status[id.train]
  X.test <- X[id.test,]
  S.test <- S[id.test,]
  time.test <- time[id.test]
  status.test <- status[id.test]
  
  # Define a sequence of alpha values to test
  alpha_values <- seq(0, 1, by = 0.1)
  # lambdas <- 10^seq(-3,3,length=100) 
  
  # Perform cross-validation for each alpha value
  cv_results <- lapply(alpha_values, function(alpha) {
    cv.glmnet(as.matrix(X.train), S.train, family = "cox", alpha = alpha,type.measure = "deviance")
  })
  
  # Extract the mean cross-validated error for each alpha
  mean_cv_errors <- sapply(cv_results, function(cv) min(cv$cvm))
  
  # Find the alpha with the minimum cross-validated error
  optimal_alpha <- alpha_values[which.min(mean_cv_errors)]
  print(paste0("optimal_alpha",optimal_alpha))
  
  pen.lik <- function(alpha)
  {
    cv.pen <- try(cv.glmnet(as.matrix(X.train), S.train, family = "cox", alpha=alpha))
    print(paste0("alpha=",alpha))
    #genes selected with optimal model fitting
    coef.min <-  coef(cv.pen, s = "lambda.min")
    active.min <- which(coef.min != 0)
    index.min <- coef.min[active.min]
    sel.pen <- rownames(coef.min)[active.min]
    #risk score
    rs.train <- predict(cv.pen,newx=as.matrix(X.train),s="lambda.min")
    rs.test <- predict(cv.pen,newx=as.matrix(X.test),s="lambda.min")
    #lasso fit
    coxfit <- glmnet(X.train, S.train, family = "cox", alpha=alpha)
    return(list(cv.pen=cv.pen,coxfit=coxfit,rs.train=rs.train,rs.test=rs.test,sel.pen=sel.pen))
  }
  
  #elastic net method with all genes
  pen.elanet <- pen.lik(optimal_alpha)
  if (length(pen.elanet$sel.pen)==0) next
  cat(i)
  
  #median stratification
  good.prog <- (pen.elanet$rs.test < median(pen.elanet$rs.train))
  fit <- survfit(S.test ~ good.prog)
  # print(paste0("fit",fit))
  # fit.all[i] <- fit
  
  
  #logrank test for test set
  logrank <- survdiff(S.test ~ good.prog)
  p.logrank <- 1-pchisq(logrank$chisq, 1)
  
  p.logrank.all[i] <- p.logrank
  
  #cox regression on risk score
  pi <- coxph(S.test ~ pen.elanet$rs.test, method = "breslow")
  p.pi <- 1-pchisq(pi$score, 1)
  print(paste0("p.logrank",p.logrank))
  
  if (p.logrank < 0.05){
    ## Plot KM
    #kaplan-meier curves
    
    # print(ggsurvplot(fit,data = PAAD_meth_match_mat,
    #            title = ("Kaplan-Meier Curve"),
    #            pval = TRUE, conf.int = TRUE,
    #            risk.table = TRUE, # Add risk table
    #            risk.table.col = "strata", # Change risk table color by groups
    #            linetype = "strata", # Change line type by groups
    #            surv.median.line = "hv", # Specify median survival
    #            ggtheme = theme_bw(),
    #            palette = c("#990000", "#000099")))
    # dev.off()
    
    plot(fit, lwd = 2, lty = c(1,1), col = c("red","blue"), xlab = 'Time', ylab = 'Estimated Survival Function')
    legend("topright", legend=c('high risk', 'low risk'), lty = c(1,1), col = c("red", "blue"), lwd = 2)
    text(1,0.05,paste("p=",round(p.logrank,4)))
    title("Kaplan-Meier Curve")
    
    print("Median OS:") 
    print(paste("high risk patient group:", round(unname(summary(fit)$table[,'median']*12)[1],2), "months"))
    print(paste("low risk patient group:", round(unname(summary(fit)$table[,'median']*12)[2],2), "months"))
    print("")
    print("")
    print(pen.elanet$sel.pen)
  }
}

```
