---
title: "Pathway analysis of RNA-seq(PAAD) differential expressed genes data from TCGA"
output: html_document
author:
- name: Aditya Pawar
date: "2024-06-12"
---

#### This script uses RNA-seq differentially expressed data for PAAD cancer from TCGA analysed using DESeq2. The differentially expressed genes are filtered on panel of genes and mapped to different public pathway databases.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the dependent libraries required
```{r library, warning=FALSE,message=FALSE}
library(org.Hs.eg.db)
library(msigdbr)
library(fgsea)
library(tidyr)
library(biomaRt)
library(dplyr)
library(data.table)
library(readr)
```

Functions useful for downstream pathway analysis
```{r pathway functions}
fora <- function(pathways, genes, universe, minSize=1, maxSize=length(universe)-1) {
  # Error if pathways is not a list
  if (!is.list(pathways)) {
    stop("pathways should be a list with each element containing genes from the universe")
  }
  
  # Warning message for duplicate gene names
  if (any(duplicated(universe))) {
    warning("There were duplicate genes in universe, they were collapsed")
    universe <- unique(universe)
  }
  
  minSize <- max(minSize, 1)
  
  pathwaysFiltered <- lapply(pathways, function(p) { unique(na.omit(match(p, universe))) })
  pathwaysSizes <- sapply(pathwaysFiltered, length)
  
  toKeep <- which(minSize <= pathwaysSizes & pathwaysSizes <= maxSize)
  
  if (length(toKeep) == 0){
    return(data.table(pathway=character(),
                      pval=numeric(),
                      padj=numeric(),
                      overlap=integer(),
                      size=integer(),
                      overlapGenes=list()))
  }
  
  pathwaysFiltered <- pathwaysFiltered[toKeep]
  pathwaysSizes <- pathwaysSizes[toKeep]
  
  
  if (!all(genes %in% universe)) {
    warning("Not all of the input genes belong to the universe, such genes were removed")
  }
  genesFiltered <- unique(na.omit(match(genes, universe)))
  
  overlaps <- lapply(pathwaysFiltered, intersect, genesFiltered)
  
  overlapGenes <- lapply(overlaps, function(x) universe[x])
  
  overlapsT <- data.table(
    q=sapply(overlaps, length),
    m=sapply(pathwaysFiltered, length),
    n=length(universe)-sapply(pathwaysFiltered, length),
    k=length(genesFiltered))
  
  # q-1 because we want probability of having >=q white balls
  pathways.pvals <- with(overlapsT,
                         phyper(q-1, m, n, k, lower.tail = FALSE))
  
  res <- data.table(pathway=names(pathwaysFiltered),
                    pval=pathways.pvals,
                    padj=p.adjust(pathways.pvals, method="BH"),
                    overlap=overlapsT$q,
                    size=overlapsT$m,
                    overlapGenes=overlapGenes)
  
  res <- res[order(pval),]
  res
}

collapsePathwaysORA <- function(foraRes,
                                pathways,
                                genes,
                                universe,
                                pval.threshold=0.05) {
  
  pathways <- pathways[foraRes$pathway]
  pathways <- lapply(pathways, intersect, universe)
  
  parentPathways <- setNames(rep(NA, length(pathways)), names(pathways))
  
  for (i in seq_along(pathways)) {
    p <- names(pathways)[i]
    if (!is.na(parentPathways[p])) {
      next
    }
    
    pathwaysToCheck <- setdiff(names(which(is.na(parentPathways))), p)
    
    if (length(pathwaysToCheck) == 0) {
      break
    }
    
    minPval <- setNames(rep(1, length(pathwaysToCheck)), pathwaysToCheck)
    
    u1 <- setdiff(universe, pathways[[p]])
    foraRes1 <- fora(pathways = pathways[pathwaysToCheck],
                     genes=intersect(genes, u1),
                     universe=u1,
                     maxSize=length(u1)-1)
    minPval[foraRes1$pathway] <- pmin(minPval[foraRes1$pathway], foraRes1$pval)
    
    u2 <- pathways[[p]]
    foraRes2 <- fora(pathways = pathways[pathwaysToCheck],
                     genes=intersect(genes, u2),
                     universe=u2,
                     maxSize=length(u2)-1)
    minPval[foraRes2$pathway] <- pmin(minPval[foraRes2$pathway], foraRes2$pval)
    
    parentPathways[names(which(minPval > pval.threshold))] <- p
  }
  
  return(list(mainPathways=names(which(is.na(parentPathways))),
              parentPathways=parentPathways))
}
```

Load the RNA-seq differentially expressed gene data

```{r load data}
## Read deg data
deg <- read.csv("DEG_results_Primary_Tumor vs. Solid_Tissue_Normal.csv",
                 header = TRUE, sep = ",")

names(deg)[names(deg) == "X"] <- "genes"
# Connect to the Ensembl BioMart database
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

GTF <- "gencode.v42.chr_patch_hapl_scaff.annotation.gtf"
gtf_df <- as.data.frame(rtracklayer::import(GTF))
gtf_df$gene_id <- sub("\\..*","\\", gtf_df$gene_id) # remove the ensembl numbering
tx2gene <- gtf_df %>% dplyr::select(transcript_id, gene_id) %>%
  filter(!is.na(transcript_id)) %>% 
  dplyr::distinct()

rdata <- tx2gene %>% dplyr::select(gene_id) %>% distinct() %>% 
  inner_join(gtf_df %>% dplyr::select(gene_id, gene_name), by = 'gene_id') %>%
  distinct() 
rdata$gene_name[is.na(rdata$gene_name)] <- rdata$gene_id[is.na(rdata$gene_name)]

deg_data <- merge(deg,rdata, by.x = "genes", by.y="gene_name") 
head(deg_data)
```

#### Match the gene panel
```{r Match the output with Dr. Manne gene panel}
genes_PAAD <- sort(c(
  "BNIP3", "CADM1", "CLDN5", "DISP3", "EBF3", "EYA4", "FOXE1", "NXPH1", "RELN", "SLIT2", "SOX17", "ADAMTS1", "APC", "BMP3", "BNC1", "BRCA1", "CCDC181", "CCND2", "CDKN1C", "CDKN2B", "DAPK1", "EPB41L3", "ESR1", "ALKAL1", "GATA4", "GATA6", "GPC3", "HOXA1", "HOXA7", "IGF1", "LRFN5", "MAPT", "MARCHF11", "MEIS2", "MEST", "MGMT", "MIR-129-2/129", "MLH1", "MUC2", "MYOD1", "NPTX2", "ONECUT1", "CDKN2A", "PANTR1", "PCDH10", "PENK", "PGK1", "PGR", "PLAU", "PRKCB", "PROX1", "PXDN", "RARB", "RASSF1", "RB1", "RUNX3", "SEPTIN9", "SFRP1", "SFRP2", "SIX3", "SLC9A3", "SOCS1", "SPARC", "SPSB4", "CD2", "SST", "SYK", "TBR1", "TEAD1", "THBS1", "TYMS", "TRIM73", "UCHL1", "VHL", "VWC2", "YAP1", "ZNF781", "SEMA5A", "TFPI2", "MIR-1292/3P", "MIR-1292/5P", "MUC5AC", "ACIN1", "ALX4", "BNC1", "CDKN2A", "GSTP1", "HIC1", "MSX2", "MEST", "MGMT", "MLH1", "NEUROG1", "NPTX2", "PCDH10", "PENK", "POU4F1", "SEMA5A", "SEPTIN9", "SFRP1", "SFRP2", "SPSB4", "TFPI2", "TNFRSF10C", "WNT5A", "UCHL1", "GSTM1", "ONECUT2", "ASXL2", "TET1", "TSPAN1", "ABCC1", "ABCG2", "CDA", "ABCB11", "ABCB4", "ABCC10", "ABCC3", "METTL3", "DDX3X", "SLC29A1", "CADM1", "PTGES2", "HSPA5", "SNRPF", "ABCC5", "CES2", "SPARC", "BRCA1", "EGFR", "TNFSF10", "ABCB1", "MYEOV", "HMGA1", "JAG1", "LCN2", "NOTCH2", "RRM2", "TM4SF1", "RRM1", "MUC5AC", "SLC29A3", "SFN", "DPYD", "MAP2", "MUTYH", "SLC28A1", "BCL2L1", "MAP2", "ABCC2", "SLC22A3", "MLH1", "SFN", "VASH2", "SST", "DKK3", "ARID1A", "GSTM2", "ISG15", "SMARCA2", "MUC4", "HNF1A", "NT5C1A", "TYMS", "ABCB5", "ABCC11", "SLC28A2", "SLC22A7", "ZEB1", "TFAP2E", "NRP1", "SLC28A3", "MCL1", "PYCARD", "MUTYH", "SOX8", "CHFR", "SLFN11", "MAP3K7", "CXCL8", "SLFN11", "SLC22A2", "SLFN11", "ZEB1", "CTNNB1", "IGFBP3", "PROKR2", "ABCC1", "ABCC2", "ABCC3", "ABCC4", "ABCC5", "ABCC6", "CFTR", "ABCC8", "ABCC9", "ABCC10", "ABCC11", "ABCC12", "ABCC13", "USP8", "PRMT1", "RUNX1", "UBE2T", "GSDME", "TGM2", "DCK", "TPX2", "CPT1B", "ENO1", "CPT1B", "ATM", "DHX38", "RELL2", "GLO1", "LDHA", "CTPS2", "BICC1", "HSPB1", "CASC9", "TACC3", "ITGA3", "CXCL5", "APOL1", "ANGPTL4", "SLC2A1", "ALDOB", "KCNH2", "ABCB1", "TFAM", "TMEM139", "RRP9", "CD44", "IGF1R", "HSPB1", "TRIM31", "YEATS4", "EIF5A", "CCDC85A", "TGM2", "ITGB4", "GSDMC", "SLC38A5","GSTM1", "ONECUT2", "ASXL2", "TET1", "TSPAN1", "ABCC1", "ABCG2", "CDA", "ABCB11", "ABCB4", "ABCC10", "ABCC3", "METTL3", "DDX3X", "SLC29A1", "PTGES2", "HSPA5", "SNRPF", "ABCC5", "CES2", "SPARC", "BRCA1", "EGFR", "TNFSF10", "ABCB1", "MYEOV", "HMGA1", "JAG1", "LCN2", "NOTCH2", "RRM2", "TM4SF1", "RRM1", "MUC5AC", "SLC29A3", "SFN", "DPYD", "MAP2", "MUTYH", "SLC28A1", "BCL2L1", "MAP2", "ABCC2", "SLC22A3", "MLH1", "SFN", "VASH2", "BNIP3", "SST", "DKK3", "ARID1A", "GSTM2", "ISG15", "SMARCA2", "MUC4", "HNF1A", "NT5C1A", "TYMS", "ABCB5", "ABCC11", "SLC28A2", "SLC22A7", "ZEB1", "TFAP2E", "NRP1", "SLC28A3", "MCL1", "PYCARD", "MUTYH", "SOX8", "CHFR", "SLFN11", "MAP3K7", "CXCL8", "SLFN11", "SLC22A2", "SLFN11", "ZEB1", "CTNNB1", "IGFBP3", "PROKR2", "ABCC1", "ABCC2", "ABCC3", "ABCC4", "ABCC5", "ABCC6", "CFTR", "ABCC8", "ABCC9", "ABCC10", "ABCC11", "ABCC12", "ABCC13", "USP8", "PRMT1", "RUNX1", "UBE2T", "GSDME", "TGM2", "DCK", "TPX2", "CPT1B", "ENO1", "CPT1B", "ATM", "DHX38", "RELL2", "GLO1", "LDHA", "CTPS2", "BICC1", "HSPB1", "CASC9", "TACC3", "ITGA3", "CXCL5", "APOL1", "ANGPTL4", "SLC2A1", "ALDOB", "KCNH2", "ABCB1", "TFAM", "TMEM139", "RRP9", "CD44", "IGF1R", "HSPB1", "TRIM31", "YEATS4", "EIF5A", "CCDC85A", "TGM2", "ITGB4", "GSDMC", "SLC38A5"))

genes_PAAD <- unique(genes_PAAD)

deg_data <- deg_data[deg_data$genes %in% genes_PAAD, ]
```

Load the Pathway databases
```{r load pathway database}
universe=deg_data %>% 
  filter(!is.na(adj.P.Val)) %>% pull(gene_id)
mapping = AnnotationDbi::select(org.Hs.eg.db, universe, 'ENTREZID', 'ENSEMBL')

ranks_df <- deg_data %>% 
  filter(!is.na(adj.P.Val)) %>% # only if enough DE genes  
  arrange((logFC)) %>% 
  left_join(mapping, by = c("gene_id"="ENSEMBL")) %>% 
  filter(!is.na(ENTREZID) & !is.na(adj.P.Val))
ranks_df = ranks_df[!duplicated(ranks_df$gene_id),]
ranks <- ranks_df$logFC
names(ranks) <- ranks_df$gene_id

all_in_life=list(
  msigdbr(species = "human", category = "H") %>% mutate(gs_subcat="Hallmark"),
  msigdbr(species = "human", category = "C2", subcategory = "CP:REACTOME"),
  msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG"),
  msigdbr(species = "human", category = "C2", subcategory = "CP:PID"),
  msigdbr(species = "human", category = "C5", subcategory = "GO:BP"),
  msigdbr(species = "human", category = "C5", subcategory = "GO:MF"),
  msigdbr(species = "human", category = "C5", subcategory = "HPO"),
  msigdbr(species = "human", category = "C3", subcategory = "TFT:GTRD"),
  msigdbr(species = "human", category = "C6") %>% mutate(gs_subcat="Oncogenic")
)

```

```{r gsea, warning=FALSE,message=FALSE}
pathways_gsea_all = lapply(all_in_life, function(p){
  pathway = split(x = p$ensembl_gene, f = p$gs_name)
  db_name = paste(p$gs_cat[1], p$gs_subcat[1],sep=":")
  respath <- fgsea(pathways = pathway, 
                   stats = ranks,
                   minSize  = 5,
                   maxSize  = 500
                   )
  
  coll_respath = collapsePathways(respath[order(pval)][padj < 0.1], 
                                  pathway, ranks)
  as_tibble(respath[pathway %in% coll_respath$mainPathways])  %>% 
    mutate(database=db_name)
}) %>% bind_rows() %>% 
  mutate(analysis="GSEA")

gsea_tb=pathways_gsea_all %>% unnest(leadingEdge) %>%
  group_by(pathway) %>% 
  left_join(mapping, by =c("leadingEdge"="ENSEMBL")) %>% 
  dplyr::select(pathway, padj, NES, leadingEdge, analysis, database)
```

```{r ORA,  warning=FALSE,message=FALSE}

# check cutoff
ora_input = deg_data %>% filter(!is.na(adj.P.Val)) %>% pull(gene_id)
input_entrezid <- AnnotationDbi::select(org.Hs.eg.db, ora_input, 'ENSEMBL', columns = c('ENTREZID', 'SYMBOL'))

total_deg=length(unique(ora_input))/length(unique(mapping$ENTREZID))

pathways_ora_all = lapply(all_in_life, function(p){
  pathway = split(x = p$entrez_gene, f = p$gs_name)
  db_name = paste(p$gs_cat[1], p$gs_subcat[1],sep=":")
  respath <- fora(pathways = pathway, 
                  genes = unique(input_entrezid$ENTREZID),
                  universe = unique(mapping$ENTREZID),
                  minSize  = 5,
                  maxSize  = 500)
  coll_respath = collapsePathwaysORA(respath[order(pval)][padj < 0.1], 
                                     pathway, unique(input_entrezid$ENTREZID), unique(mapping$ENTREZID))
  as_tibble(respath[pathway %in% coll_respath$mainPathways])  %>% 
    mutate(database=db_name, NES=(overlap/size)/(total_deg))
}) %>% bind_rows() %>% 
  mutate(analysis="ORA")

ora_tb = pathways_ora_all %>% unnest(overlapGenes) %>%
  group_by(pathway) %>% 
  left_join(mapping, by =c("overlapGenes"="ENTREZID")) %>% 
  dplyr::select(pathway, padj, NES, ENSEMBL, analysis,
                database)

pathways_long = bind_rows(gsea_tb, ora_tb)
head(pathways_long)
```

Annotate ENSEMBLE ids with Genes

```{r annotate genes,  warning=FALSE,message=FALSE}

res_ <- dplyr::inner_join(deg_data,gtf_df%>% dplyr::select(gene_id),by = "gene_id") %>% distinct()

pathways_long_ <- dplyr::inner_join(pathways_long,gtf_df%>% dplyr::select(gene_id,gene_name),by = c("leadingEdge" = "gene_id")) %>% distinct()

head(pathways_long_)

write_csv(pathways_long_, "Pathway_genes_panel.csv")
```